Do researchers preferentially collaborate with colleagues of the same gender?

### 2018-05-27 manuscript/Manuscript_Claire.Rmd
Writing

---
title: "Do researchers preferentially collaborate with colleagues of the same gender?"
author:
- Luke Holman* and Claire Morandin&sect;
- '*\textit{luke.holman@unimelb.edu.au} \vspace{5mm}'
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
fontsize: 12pt
header-includes: 
  \usepackage{microtype} 
  \usepackage[all]{hypcap} 
  \usepackage{amsmath} 
  \usepackage{booktabs}
  \usepackage{caption} 
  \usepackage[labelfont=bf]{caption} 
  \usepackage{titling} 
  \pretitle{\begin{flushleft}\LARGE}
  \posttitle{\par\end{flushleft}\vskip 0.5em} 
  \preauthor{\begin{flushleft}\large \lineskip 0.5em} 
  \postauthor{\par\end{flushleft}} 
  \predate{\begin{flushleft}\large} 
  \postdate{\par\end{flushleft}}
  \usepackage{fancyhdr} 
  \pagestyle{fancy} 
  \fancyhead[LO,LE]{\textsl{\leftmark}} 
  \rhead[]{Gender and coauthorship} 
  \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} 
  \usepackage{amssymb} 
  \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and often feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping to redress gender imbalance in the academic workforce, increasing women's access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers co-publish with colleagues of the same gender, and show that this 'gender homophily' is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor, as predicted if mixed-gender teams produce better research. \vspace{5mm} \par\noindent \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining, Women in STEM.
---

@@

Because publishing, networking and collaboration are instrumental to scientific productivity and academic career advancement [@Lee_2005; @Wuchty_2007; @Abramo_2009; @Lariviere_2015], dozens of studies have tested for gender differences in these areas [e.g. @Long_1992; @Bozeman_2011; @Abramo_2013; @Badar_2013; references in Table S1 of @Holman_2018]. For example, studies have found that women tend to be less involved in international collaboration [@Lewison_2001; @Webster_2001; @Bozeman_2004; @Lariviere_2011; @Abramo_2013], collaborate less within their own university departments [@Webster_2001], have less prestigious collaborations [@Long_1990], and fewer collaborations in total [@Fuchs_2001]. These gender differences in collaboration practice presumably have multiple concurrent causes, which might include implicit and explicit gender bias [@Moss_2012], differential family obligations [@Reskin_1978; @Long_1990; @Wright_2003], concerns relating to sexual harassment [@jagsi_2016], and unequal access to conferences [@Martin_2014] and travel funds [@Bozeman_2004].

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008; @Cohn_2014]. Additionally, empirical studies suggest that mixed-gender [@Bear_2011; @Campbell_2013] or otherwise diverse [@Hong_2004] teams can produce better results on collaborative tasks than less diverse teams. For reasons such as these, several studies have tested for gender differences in collaboration frequency or pattern by examining the author lists of published research. To our knowledge, every published study of this question suggests that men co-publish with other men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015; @Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This pattern of assortative publishing has been termed 'gender homophily'.

However, we believe that prior studies of gender homophily were hindered by a largely unacknowledged statistical issue that we term the Wahlund effect (\autoref{wahlund_plot}), by analogy with a conceptually similar effect in population genetics [@Wahlund_1928]. The Wahlund effect makes it deceptively hard to measure gender homophily. Essentially, whenever coauthorship data are sampled from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can give the impression that authors preferentially publish with same-gendered colleagues if no gender preferences exist, or even if researchers actually prefer opposite-gendered colleagues. For example, a sample of literature containing bioinformatics and cell biology papers will probably contain an excess of mostly-male and mostly-female author lists, simply because researchers preferentially work with colleagues from the same discipline, and because the author gender ratio is more male-biased in bioinformatics than in cell biology [@Holman_2018].

![The Wahlund effect can make it appear as if authors prefer to publish with same-gendered colleagues, even if no such preference exists. Here, coloured circles represent male and female authors, and coauthors are linked with lines. Across the whole set of ten papers, there is an apparent excess of same-gender collaborations: there are six same-gender papers and only four mixed-gender papers, which is fewer than the $10\times2\times0.5\times0.5 = 5$ mixed-gender papers we would expect under the null hypothesis that authors assort randomly with respect to gender. However, within each subset, there is no evidence that authors prefer to publish with same-gendered individuals. The Wahlund effect will tend to inflate the frequency of same-sex coauthorships whenever the data is composed of two or more disconnected subsets of literature with different author gender ratios; these subsets could be research disciplines, older versus newer papers, or papers from authors in different countries. \label{wahlund_plot}](../figures/Figure 1 - Wahlund figure.pdf)

In the present study, we test whether academics tend to co-publish with same-gendered collaborators, while controlling for the Wahlund effect as strictly as possible. Our study uses a recently-published dataset describing the gender of 35.5m authors from 9.15m articles indexed on PubMed [@Holman_2018]. Holman et al. [-@Holman_2018] reported large differences in the gender ratio of authors across research disciplines, journals, countries, and across the years 2002-2016. We therefore tested for gender homophily after restricting the analysis to particular journals (i.e. research specialties), time periods, and countries.

# Methods
## The dataset
We used the dataset of PubMed author lists from Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every article indexed on PubMed, from the oldest articles up to the present, and attempting to assign a gender to each author based on their given name. Each journal was assigned to one of 107 scientific disciplines, using PubMed's journal categorisations where available. Because the present study focuses on co-authorship, all single-author papers were discarded. We also discarded all papers for which we could not determine the gender of every author with ${\ge}95\%$ certainty, in order to simplify the statistical analysis. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published either 0-1 or 10-11 years before the PubMed data were collected (i.e. 20^th^ August 2016). Lastly, we excluded journals with fewer than 50 suitable papers.

@@

## Calculating $\alpha$, the coefficient of homophily
Inspired by Bergstrom et al. (http://www.eigenfactor.org/gender/assortativity/measuring_homophily.pdf), we defined the coefficient of homophily as $\alpha = p - q$, where $p$ is the probability that a randomly-chosen co-author of a _male_ author is a man and $q$ is the probability that a randomly-chosen co-author of a _female_ author is a man. Therefore, $\alpha > 0$ suggests that same-gender authors publish together more often than expected if authors assort randomly with respect to gender (homophily), $\alpha < 0$ suggests that opposite-gender authors publish together more often than expected (heterophily), and $\alpha = 0$ suggests random assortment with respect to gender.

@@

## Statistical analysis
Previous authors [e.g. @Bonham_2017] have hypothesised that senior scientists preferentially recruit staff and students of the same gender, and/or that junior researchers preferentially select same-gendered mentors. In the majority of disciplines, authorship conventions mean that the first-listed author is often an early-career researcher, while the author listed last is more likely to be a senior researcher leading a research team [@Wren_2007]. Assuming that senior researchers are the main drivers of homophily and that there are enough papers with three or more authors, we predict that the last author's gender will be the strongest predictor of the remaining authors' genders (i.e. the gender of the last author will be more salient than that of the first author, or any other authorship position). This is because the first author's gender would simply be an imperfect correlate of the true causal effect, while the last author's gender would be the causal effect itself.

@@

## Theoretical expectations for $\alpha$ when the gender ratio differs between career stages
In many STEMM subjects, the gender ratio is more skewed among established researchers relative to early-career researchers [@Shaw_2012; @Holman_2018]. We hypothesised that this skew could potentially create both Wahlund effects and 'reverse' Wahlund effects. For example, imagine that the majority of collaborations are between students and professors, and that the gender ratio differs between career stages: we will then see an excess of mixed-gender coauthorships (heterophily, $\alpha < 0$), even if gender has no causal effect on the selection of coauthors. Similarly, a hypothetical field in which students work only with students, and professors with professors, would have apparent gender homophily ($\alpha > 0$).

We can think of no tractable method of controlling for this issue using our dataset, which contains no information on career stage. Therefore, we instead decided to derive the theoretical expectations for $\alpha$ when there is a difference in gender ratio across career stages, in order to determine whether this effect could appreciably affect our conclusions. For simplicity, our calculations assume there are only two career stages, though the conclusions are likely to generalise to a realistic, multi-tiered career ladder. Under the null model that gender has no causal effect on collaboration, we calculated $\alpha$ for various combinations of the four free parameters, i.e. the gender ratios for early- and late-career researchers, and the relative frequency of collaborations between early-early, early-late, and late-late collaborations. We then used the theoretical expectations for $\alpha$ to qualify our main conclusions (see Results). The Online Supplementary Material gives annotated R code used to derive the theoretical expectations.

@@

$\alpha'$ was significantly higher in the literature sample from 2015-16 relative to 2005-6, but the difference in means was small (Figure S3; Effect of the fixed factor 'Time period' in a linear mixed model of the data for all author positions: Cohen's $d$ = $0.09{\pm}0.04$, $t_{948}$ = 2.51, p = 0.012).

@@

## Theoretical expectations for $\alpha$ when the gender ratio differs between career stages
As shown in \autoref{simulation}, $\alpha$ is expected to be non-zero, even if collaborators are randomly selected with respect to gender, provided that there is a gender gap between career stages. The extent to which $\alpha$ deviates from zero depends on the relative frequencies of collaboration within and between career stages. When >50% of collaborations were between early and established researchers, we expect gender heterophily ($\alpha < 0$). Conversely, when >50% of collaborations occured within career stages, we expect gender homophily ($\alpha > 0$). In a few parameter spaces (shown in red; \autoref{simulation}), $\alpha$ was quite high, and overlapped with the values that we estimated (\autoref{alpha_histograms}).

@@

Despite this overlap, \autoref{simulation} suggests that our main conclusions (and those of other studies of gender homophily) are probably robust to this career stage issue. We only expect strongly positive $\alpha$ when A) the gender ratio is highly skewed across career stages (e.g. a 10-fold difference), and B) collaborations between early and established researchers are very rare (e.g. <10% of the total). Both of these conditions are unlikely: the gender gap across careers stages is generally less than 10-fold [@Shaw_2012; @Holman_2018], and it is very common for early-career researchers to co-publish with an established mentor [@macaluso_2016]. However, one can get $\alpha > 0$ for realistic combinations of parameters, e.g. a moderate shortage of women in senior positions coupled with a moderate excess of within-career stage collaboration, suggesting this effect might contribute to some of the observed homophily.

Lastly, we note that if there is a gender gap between career stages and coauthorships between early-career and established researchers comprise ${\ge}50\%$ of the total, then the baseline expectation for $\alpha$ is actually less than zero (blue areas in \autoref{simulation}). Therefore, our results might under-estimate the extent to which researchers preferentially select same-gendered collaborators in some cases.

# Discussion
We found evidence that researchers publish with same-gendered coauthors more often than expected under random assortment, even after implementing stringent controls for 'spurious' gender homophily resulting from Wahlund effects (\autoref{wahlund_plot}). Our study therefore reaffirms several earlier results and establishs their generality across the life sciences [@Ferber_1980; @McDowell_1992; @Bentley_2003; @Crow_2015; @Ghiasi_2015; @Araujo_2017a; @Fahmy_2017; @Jadidi_2017; @Teele_2017; @Zettler_2017]. Relatively few journals had $\alpha'$ values below zero, and almost no journals showed statistically significant gender heterophily after controlling for multiple testing. The excess of same-gender coauthorships was quite strong: many journals had $\alpha' > 0.1$, indicating that the gender ratio of men's and women's coauthors differs by >10% in absolute terms. In relative terms, this effect appears even more striking: for example, if the proportion of male coauthors is 80% for men and 70% for women ($\alpha' = 0.1$), the results imply that women publish with women 50% more often than men do.

An important limitation of our study is that we cannot reliably ascertain the cause(s) of the observed excess of same-gender coauthorships. As well as conscious or unconcious selection of collaborators based on gender, our results could be partly or wholly explained by uncontrolled Wahlund effects. We suspect the contribution of these to be minor, for four reasons: we found positive $\alpha'$ after controlling for three obvious sources of Wahlund effect; there was no inflation of $\alpha'$ in highly multidisciplinary journals; restricting the data by country yielded similar estimates of $\alpha'$; and we showed that differences in gender ratio between career stages are unlikely to fully explain our results. On balance, we believe the data suggest that researchers preferentially select same-gendered collaborators, the magnitude of this effect is uncertain.

Some disciplines exhibit especially skewed gender ratios; e.g. most authors in physics and computer science are men, while most authors in nursing are women [@Holman_2018]. We hypothesised that these gender-imbalanced disciplines might show the strongest gender homophily, e.g. because being in the minority might increase one's drive to seek out same-gendered colleagues. Contrary to this hypothesis, we found no evidence that gender homophily is restricted to particular disciplines: $\alpha'$ was similarly high across the board \autoref{alpha_histograms}. Moreover, $\alpha'$ was lower for journals with a male-biased author gender ratio relative to those with an even gender ratio. Interestingly, $\alpha'$ was highest in journals with a female-biased author gender ratio. One interpretation of this result is that men are more likely to preferentially seek out male collaborators in fields where men are a minority, relative to the homophily displayed by women in fields where women are a minority. However, this latter result is only tentatively supported since there are so few journals on PubMed in which most authors are women \autoref{alpha_gender_ratio}.

@@

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or bullying from men [@jagsi_2016], which is clearly very concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (i.e. recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern because the results might reflect discrimination against women by established scientists [@Moss_2012], or avoidance by women of elite research groups (e.g. due to 'imposter syndrome' or the perception that these groups are hostile to women). We also found a little evidence that gender homophily is deterimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], then this result provides non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Campbell_2013].

Homophily might also have more benign causes. Collaboration is often especially enjoyable when working with like-minded people, who might be same-gendered more often than not. We also suppose that some people -- women especially -- consciously choose to preferentially collaborate with women whenever multiple qualified candidates are available, in order to help close the gender gap in the workforce. Related to this, Ghiasi et al. [-@Ghiasi_2015] wrote that women who mostly work with men are 'compliant [in reproducing] male-dominated scientific structures', although we think this charge is unfair. We believe that everyone (not only women) should take steps to ensure that they are not overlooking or excluding female colleagues, particularly since men are more likely to occupy stable, senior jobs that afford them the luxury of choice.

>>>>

---
title: "Do researchers preferentially collaborate with colleagues of the same gender?"
author:
- Luke Holman* and Claire Morandin&sect;
- '*\textit{luke.holman@unimelb.edu.au} \vspace{5mm}'
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
fontsize: 12pt
header-includes: 
  \usepackage{microtype} 
  \usepackage[all]{hypcap} 
  \usepackage{amsmath} 
  \usepackage{booktabs}
  \usepackage{caption} 
  \usepackage[labelfont=bf]{caption} 
  \usepackage{titling} 
  \pretitle{\begin{flushleft}\LARGE}
  \posttitle{\par\end{flushleft}\vskip 0.5em} 
  \preauthor{\begin{flushleft}\large \lineskip 0.5em} 
  \postauthor{\par\end{flushleft}} 
  \predate{\begin{flushleft}\large} 
  \postdate{\par\end{flushleft}}
  \usepackage{fancyhdr} 
  \pagestyle{fancy} 
  \fancyhead[LO,LE]{\textsl{\leftmark}} 
  \rhead[]{Gender and coauthorship} 
  \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} 
  \usepackage{amssymb} 
  \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and often feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping to redress gender imbalance in the academic workforce, increasing women's access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers co-publish with colleagues of the same gender, and show that this 'gender homophily' is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor, as predicted if mixed-gender teams produce better research. We discuss potential causes of gender homophily in science. \vspace{5mm} \par\noindent \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining, Women in STEM.
---

@@

Because publishing, networking and collaboration are instrumental to scientific productivity and academic career advancement [@Lee_2005; @Wuchty_2007; @Abramo_2009; @Lariviere_2015], dozens of studies have tested for gender differences in these areas [e.g. @Long_1992; @Bozeman_2011; @Abramo_2013; @Badar_2013; references in Table S1 of @Holman_2018]. For example, studies have found that women tend to be less involved in international collaboration [@Lewison_2001; @Webster_2001; @Bozeman_2004; @Lariviere_2011; @Abramo_2013], collaborate less within their own university departments [@Webster_2001], have less prestigious collaborations [@Long_1990], and fewer collaborations in total [@Fuchs_2001]. These gender differences in collaboration practice presumably have multiple causes, which might include implicit and explicit gender bias [@Moss_2012], differential family obligations [@Reskin_1978; @Long_1990; @Wright_2003], concerns relating to sexual harassment [@jagsi_2016], and unequal access to conferences [@Martin_2014] and travel funds [@Bozeman_2004].

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008; @Cohn_2014]. Additionally, empirical studies imply that mixed-gender [@Bear_2011; @Campbell_2013] or otherwise diverse [@Hong_2004] teams produce better results on collaborative tasks than less diverse teams. For reasons such as these, several studies have tested for gender differences in collaboration frequency or pattern by examining the author lists of published research. To our knowledge, every published study of this question suggests that men co-publish with other men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015; @Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This pattern of assortative publishing has been termed 'gender homophily'.

However, we believe that prior studies of gender homophily were hindered by a largely unacknowledged statistical issue that we term the Wahlund effect (\autoref{wahlund_plot}), by analogy with a conceptually similar effect in population genetics [@Wahlund_1928]. The Wahlund effect makes it deceptively difficult to infer gender-based preferences simply by counting same- and mixed-gender coauthorships. Essentially, whenever coauthorship data are sampled from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can give the impression that authors preferentially publish with same-gendered colleagues if no gender preferences exist, or even if opposite-gendered colleagues are prefered. For example, a sample of literature containing bioinformatics and cell biology papers will probably contain an excess of mostly-male and mostly-female author lists, simply because researchers preferentially work with colleagues from the same discipline, and because the author gender ratio is more male-biased in bioinformatics than in cell biology [@Holman_2018].

![The Wahlund effect can make it appear as if authors prefer to publish with same-gendered colleagues, even if no such preference exists. Here, coloured circles represent male and female authors, and coauthors are linked with lines. Across the whole set of ten papers, there is an apparent excess of same-gender collaborations: there are six same-gender papers and only four mixed-gender papers, which is fewer than the $10\times2\times0.5\times0.5 = 5$ mixed-gender papers expected under the null hypothesis that authors assort randomly with respect to gender. However, within each subset, there is no evidence that authors prefer to publish with same-gendered individuals (if anything, this small dataset suggests gender heterophily). The Wahlund effect will tend to inflate the frequency of same-sex coauthorships whenever the data is composed of two or more disconnected subsets of literature with different author gender ratios; these subsets could be research disciplines, older versus newer papers, or papers from authors in different countries. \label{wahlund_plot}](../figures/Figure 1 - Wahlund figure.pdf)

In the present study, we test whether researchers working in the life sciences tend to co-publish with same-gendered collaborators, while controlling for the Wahlund effect as strictly as possible. Our study uses a recently-published dataset describing the gender of 35.5m authors from 9.15m articles indexed on PubMed [@Holman_2018]. Holman et al. [-@Holman_2018] reported large differences in the gender ratio of authors across research disciplines, journals, countries, and across the years 2002-2016. We therefore tested for gender homophily after restricting the analysis to particular journals (i.e. research specialties), time periods, and countries.

# Methods
## The dataset
We used the dataset of PubMed author lists from Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every article indexed on PubMed and attempting to assign a gender to each author based on their given name. Each journal was assigned to one of 107 scientific disciplines, using PubMed's journal categorisations where available. Because the present study focuses on co-authorship, all single-author papers were discarded. We also discarded all papers for which we could not determine the gender of every author with ${\ge}95\%$ certainty, in order to simplify the statistical analysis. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published either 0-1 or 10-11 years before the PubMed data were collected (i.e. 20^th^ August 2016). Lastly, we excluded journals with fewer than 50 suitable papers.

@@

## Calculating $\alpha$, the coefficient of homophily
Following Bergstrom et al. [-@bergstrom_2016], we defined the coefficient of homophily as $\alpha = p - q$, where $p$ is the probability that a randomly-chosen co-author of a _male_ author is a man and $q$ is the probability that a randomly-chosen co-author of a _female_ author is a man. Therefore, $\alpha > 0$ suggests that same-gender authors publish together more often than expected if authors assort randomly with respect to gender (homophily), $\alpha < 0$ suggests that opposite-gender authors publish together more often than expected (heterophily), and $\alpha = 0$ suggests random assortment with respect to gender. Like the Wahlund effect, $\alpha$ is borrowed from population genetics; for a set of 2-author papers, it is equivalent to the coefficient of inbreeding [@wright_1949]. Mathematical work illustrates that $\alpha$ is closely related to network-based alternative methods for quantifying homophily [@wang_2016].

@@

## Statistical analysis
Previous authors [e.g. @sheltzer_2014; @Bonham_2017] have hypothesised that senior scientists preferentially recruit staff and students of the same gender, and/or that junior researchers preferentially select same-gendered mentors. In the majority of PubMed-indexed disciplines, authorship conventions mean that the first-listed author is often an early-career researcher, while the author listed last is more likely to be a senior researcher leading a research team [@Wren_2007]. Assuming that senior researchers are the main drivers of homophily and that there are enough papers with three or more authors, we predict that the last author's gender will be the strongest predictor of the remaining authors' genders (i.e. the gender of the last author will be more salient than that of the first author, or any other authorship position). This is because the first author's gender would simply be an imperfect correlate of the true causal effect, while the last author's gender would be the causal effect itself.

@@

## Theoretical expectations for $\alpha$ when the gender ratio differs between career stages
In many STEMM subjects, the gender ratio is more skewed among established researchers relative to early-career researchers [@Shaw_2012; @Holman_2018]. We hypothesised that this skew could potentially create both Wahlund effects and 'reverse' Wahlund effects. For example, imagine that the majority of collaborations are between students and professors, and that the gender ratio differs between career stages: we will then see an excess of mixed-gender coauthorships (heterophily, $\alpha < 0$), even if gender has no direct, causal effect on coauthor selection. Similarly, a hypothetical field in which students work only with students, and professors with professors, would have apparent gender homophily ($\alpha > 0$).

We can think of no tractable method of controlling for this issue using our dataset, which contains no information on career stage. Therefore, we instead decided to derive the theoretical expectations for $\alpha$ when there is a difference in gender ratio across career stages, in order to determine if and how this effect should change our inferences. For simplicity, our calculations assume there are only two career stages, though the conclusions probably generalise to a realistic, multi-tier career ladder. Under the null model that gender has no causal effect on collaboration, we calculated $\alpha$ for various combinations of the four free parameters, i.e. the gender ratios for early- and late-career researchers, and the relative frequency of collaborations between early-early, early-late, and late-late collaborations. We then used the theoretical expectations for $\alpha$ to qualify our main conclusions (see Results). The Online Supplementary Material gives annotated R code used to derive the theoretical expectations.

@@

$\alpha'$ was significantly higher in the literature sample from 2015-16 relative to 2005-6, though the difference in means was small (Figure S3; Effect of the fixed factor 'Time period' in a linear mixed model of the data for all author positions: Cohen's $d$ = $0.09{\pm}0.04$, $t_{948}$ = 2.51, p = 0.012).

@@

## Theoretical expectations for $\alpha$ when the gender ratio differs between career stages
As shown in \autoref{simulation}, we predict that $\alpha$ is expected to be non-zero, even if collaborators are randomly selected with respect to gender, provided that there is a gender gap between career stages. The extent to which $\alpha$ deviates from zero depends on the relative frequencies of collaboration within and between career stages. When >50% of collaborations were between early and established researchers, we expect gender heterophily ($\alpha < 0$). Conversely, when >50% of collaborations occured within career stages, we expect gender homophily ($\alpha > 0$). In a few parameter spaces (shown in red; \autoref{simulation}), $\alpha$ was quite high, and overlapped with the values that we estimated (\autoref{alpha_histograms}).

@@

Despite this overlap, \autoref{simulation} suggests that our main conclusions (and those of other studies of gender homophily) are probably robust to this career stage issue. We only expect strongly positive $\alpha$ when A) the gender ratio is highly skewed across career stages (e.g. a 5-fold difference), and B) collaborations between early and established researchers are very rare (e.g. <10% of the total). Both of these conditions are untrue for most fields: the gender gap across careers stages is generally less pronounced [e.g. @Shaw_2012; @Holman_2018], and it is very common for early-career researchers to co-publish with an established mentor [@macaluso_2016]. However, one can get $\alpha > 0$ for realistic combinations of parameters, e.g. a moderate shortage of women in senior positions coupled with a moderate excess of within-career stage collaboration, suggesting this effect might contribute to some of the observed homophily.

Lastly, we note that if there is a gender gap between career stages and coauthorships between early-career and established researchers comprise >50% of the total, then the baseline expectation for $\alpha$ is actually less than zero (blue areas in \autoref{simulation}). Therefore, our results might under-estimate the extent to which researchers preferentially select same-gendered collaborators in some cases.

# Discussion
We found evidence that researchers publish with same-gendered coauthors more often than expected under random assortment, even after implementing stringent controls for Wahlund effects (\autoref{wahlund_plot}). Our study therefore reaffirms earlier studies' conclusions [e.g. @Ferber_1980; @McDowell_1992; @Bentley_2003; @Crow_2015; @Ghiasi_2015; @Araujo_2017a; @Fahmy_2017; @Jadidi_2017; @Teele_2017; @Zettler_2017] and establishes their generality across the life sciences. Relatively few journals had $\alpha'$ values below zero, and almost no journals showed statistically significant gender heterophily after controlling for multiple testing. The excess of same-gender coauthorships was quite strong: many journals had $\alpha' > 0.1$, indicating that the gender ratio of men's and women's coauthors differs by >10% in absolute terms. In relative terms, this effect appears even more striking: for example, if the proportion of male coauthors is 80% for men and 70% for women ($\alpha' = 0.1$), the results imply that women publish with women 50% more often than men do.

An important limitation of our study is that we cannot reliably ascertain the cause(s) of the observed excess of same-gender coauthorships. As well as conscious or unconcious selection of collaborators based on gender, our results could be partly or wholly explained by uncontrolled Wahlund effects. However, we suspect the contribution of these to be minor, for four reasons: we found positive $\alpha'$ after controlling for three obvious sources of Wahlund effect; there was no inflation of $\alpha'$ in highly multidisciplinary journals; restricting the data by country yielded similar estimates of $\alpha'$; and we showed that differences in gender ratio between career stages are unlikely to fully explain our results. On balance, we believe the data provide good evidence that researchers preferentially select same-gendered collaborators, although the effect size of this preference is uncertain.

Some disciplines exhibit especially skewed gender ratios; e.g. most authors in physics and computer science are men, while most authors in nursing are women [@Holman_2018]. We hypothesised that these gender-imbalanced disciplines might show the strongest gender homophily, e.g. because being in the minority might increase one's drive to seek out same-gendered colleagues. Contrary to this hypothesis, we found no evidence that gender homophily is restricted to particular disciplines: $\alpha'$ was similarly high across the board \autoref{alpha_histograms}. Moreover, $\alpha'$ was lower for journals with a male-biased author gender ratio relative to those with an even gender ratio. Interestingly, $\alpha'$ was highest in journals with a female-biased author gender ratio. One interpretation of this result is that men are more likely to preferentially seek out male collaborators in fields where men are a minority, relative to the homophily displayed by women in fields where women are a minority. However, this latter result is only tentatively supported since there are so few journals in which most authors are women \autoref{alpha_gender_ratio}.

@@

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment, sexism or microaggressions from men [@jagsi_2016], which is clearly very concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since Sheltzer and Smith [-@sheltzer_2014]'s results might reflect discrimination against women [@Moss_2012], or avoidance by women of elite research groups (e.g. due to 'imposter syndrome' or the perception that these groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], then this result provides non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Campbell_2013].

Homophily might also have more benign causes. Collaboration is often most enjoyable and productive when working with like-minded people, who might be same-gendered more often than not. We also suppose that some people consciously choose to preferentially collaborate with women in order to help close the gender gap in the workforce; this would create homophily if this motivation applies to more women than men. Related to this, Ghiasi et al. [-@Ghiasi_2015] concluded that women in engineering are 'compliant [in reproducing] male-dominated scientific structures' because they do not collaborate often enough with other women (coauthorships involving two women were about 30% more common than expected under the null). In contrast, we believe that everyone (not just women) should work to ensure that they are not overlooking or excluding female colleagues, or devoting less effort to collaborations with them. Men are more often well placed to help in this regard, given their overrepresentation in jobs that allow the hiring of staff.
